---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

```{r echo=TRUE, results='markup', message=F, warning=F}
library(dplyr)
library(ggplot2)

unzip("activity.zip")
d <- read.csv("activity.csv", header=TRUE, na.strings = "NA")
d$date <- as.Date(d$date)
```


## What is mean total number of steps taken per day?
First, We will group the data frame by date.
```{r echo=TRUE, results='markup',message=F, warning=F}
# group by date
dGroupedByDate <- group_by(d, date)
# calculate mean, median, and totalsteps using the summarise function
dSummary<- summarise(dGroupedByDate, meanSteps=mean(steps), medianSteps=median(steps, na.rm=TRUE),  totalSteps=sum(steps))
```


**Histogram of the total number of steps taken each day**
let's plot the data to see the total number of steps taken each day
```{r echo=TRUE, results='markup',message=F, warning=F}
qplot(totalSteps, data=dSummary,geom="histogram", main="Histogram of the total number of steps taken each day")
```

**Mean and median total number of steps taken per day:**
```{r echo=TRUE, results='markup',message=F, warning=F}
print(dSummary, n=nrow(dSummary))
```


## What is the average daily activity pattern?
First, we need to group the observations by time intervals accross all days. then, calaculate the mean of steps per each interval.
```{r echo=TRUE, results='markup',message=F, warning=F}
dGroupedByInterval <- group_by(d, interval)
dSummaryInterval<- summarise(dGroupedByInterval, meanSteps=mean(steps, na.rm=TRUE))

qplot(interval, meanSteps, data=dSummaryInterval,geom="line")
dSummaryInterval<- arrange(dSummaryInterval, desc(meanSteps)) # sort descending by mean Steps
intervalWithMaxSteps <- dSummaryInterval[1,1] 
```
**The 5 min interval with max average steps is: `r intervalWithMaxSteps`**

## Imputing missing values
**Total number of missing values in the dataset (i.e. the total number of rows with NAs)**
```{r echo=TRUE, results='markup',message=F, warning=F}
NAStepsCount <- summarise(d, NAcount=sum(is.na(steps)), total.count=n())
percentageOfNA <- (NAStepsCount$NAcount[1]/NAStepsCount$total.count[1])*100
```
**Total number of NA: `r NAStepsCount$NAcount[1]`**

Percentage of NA among observations: `r percentageOfNA` %

**Strategy to Fill NA enteries**
We can fill the NA enteries with the averages of their corresponding 5-min intervals accross all the days. If the avergae is not avaiable (i.e. incase this interval had no observation recorded accross all days, we will set it to zero)
```{r echo=TRUE, results='markup',message=F, warning=F}
# Copying original dataset to a new data frame before replacing the NAs
dEstimated <- d
# creating a lookup list for the average of each interval
# note the interval will be a chanracter index for the generated array
intervalMean <-tapply(d$steps, d$interval,mean, na.rm=TRUE)
#loop on all rows with NA enteries and replace it with the interval average
for (i in which(is.na(dEstimated)))
  {
  avgSteps <- intervalMean[as.character(dEstimated[i,3])]
  dEstimated[i,1] <- ifelse(is.na(avgSteps), 0, avgSteps)
  }
```

**Histogram of the total number of steps taken each day (after replacing NAs with average steps of same intervals)**
```{r echo=TRUE, results='markup',message=F, warning=F}
dEstimatedGroupedByDate <- group_by(dEstimated, date)
dEstimatedSummary<- summarise(dEstimatedGroupedByDate, meanSteps=mean(steps), medianSteps=median(steps, na.rm=TRUE),  totalSteps=sum(steps))
qplot(totalSteps, data=dEstimatedSummary,geom="histogram")

```


**Mean and median total number of steps taken per dayay (after replacing NAs with average):**
```{r echo=TRUE, results='markup',message=F, warning=F}
print(dEstimatedSummary, n=nrow(dEstimatedSummary))
```

**Do these values differ from the estimates from the first part of the assignment?**
let's have a quick look on the summary of the original and imputed table
```{r echo=TRUE, results='markup',message=F, warning=F}
summary(dSummary)
summary(dEstimatedSummary)
colSums(dSummary[,2:4], na.rm = T)
colSums(dEstimatedSummary[,2:4], na.rm = T)
```
By comparing the data summaries above, the mean, median and average total steps have the following chnages:
- The Total average steps has increased after adding the imputated/estimated data
- The overall mean of steps (for all days) have remained the same, however it has chnaged for the days with imputated/estimates data.
- The median has chnaged for the days with estimated enteries.

**What is the impact of imputing missing data on the estimates of the total daily number of steps?**

```{r echo=TRUE, results='markup',message=F, warning=F}
originalTotalDays <- summarise(dSummary, total.steps=sum(totalSteps, na.rm = T))
imputedTotalDays <- summarise(dEstimatedSummary, total.steps=sum(totalSteps, na.rm = T))

totDaysDiff <- imputedTotalDays[[1]] -originalTotalDays[[1]]
totDaysDiffPercentage <- (totDaysDiff/originalTotalDays) * 100
```
The imputed days added `r totDaysDiff` extra steps representing ad addition of `r totDaysDiffPercentage` % of the original total steps.


## Are there differences in activity patterns between weekdays and weekends?
Creating a new factor variable (dayType) in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day

```{r echo=TRUE, results='markup',message=F, warning=F}
dEstimated$dayType <- ifelse((weekdays(dEstimated$date) %in% c('Sunday','Saturday')), 'weekend', 'weekday')
dEstimated$dayType <- as.factor(dEstimated$dayType)
str(dEstimated)
```

a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was created using simulated data

```{r echo=TRUE, results='markup',message=F, warning=F}
dGrpByDayTypeInterval <- group_by(dEstimated, dayType, interval)
dSummaryDayTypeInterval<- summarise(dGrpByDayTypeInterval, meanSteps=mean(steps, na.rm=TRUE))
qplot(interval, meanSteps, data=dSummaryDayTypeInterval,geom="line", facets = dayType~., ylab = "Average Number of Steps")
```


